A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-09 tidyindex.qmd
re-write the index definition section and shift of research direction in intro

---
title: "Demo arXiv template"
format:
  tandf-pdf:
    keep-tex: true  
pdf-engine: pdflatex
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
    url: https://huizezhangsh.netlify.app/
  - name: Collaborators
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
abstract: |
  - indexes, useful, quantify severity, early monitoring, 
  - A huge number of indexes have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
  - One of the reasons for this is the plenty of indexes are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
  - The paper describes a general pipeline framework to construct indexes from spatio-temporal data, 
  - This allows all the indexes to be constructed through a uniform data pipeline and different indexes to vary on the details of each step in the data pipeline and their orders. 
  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indexes into small building blocks shared by most of the indexes.
  - The framework will be demonstrated with drought indexes as examples, but appliable in general to environmental indexes constructed from multivariate spatio-temporal data
keywords:
  - indexes
  - data pipeline
  - software design
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: console
---

@@

@@

  - Indexes can be composed from single variables or multiple variables (composite index)
  - to make the definition even more complex is that indexes themselves are used with other variables to compose indexes (many drought indexes using an index called SPI). 
  - Some indexes may be more simple than you think:  
    - a single transformation (NDVI [@tucker_red_1979] takes two spectrum: red and near-infrared and calculates as $(\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})$)
    - indexes sometimes are similar to features extracted from time series data: ETCCDI climate change indices [@climate-indexes]: SU - number of summer days: annual count of days when TX > 25 degree
  - The index pipeline and software implementation we propose are applicable to all indexes mentioned above, while the main focus of this paper is on indexes that combine multiple variables/indexes since more skills are required to diagnose those indexes to understand how they behave.

@@

  - Difficult to reach consensus on composite/multivariate indexes. 
  - Given the complexity of multivariate indexes as compared to univariate indexes, it raises more doubt to persuade people to use it. 
  - Typical ways to evaluate new indexes is to compare it with the standard indexes (i.e. SPI in drought index) or a response variable (i.e. crop yield for monitoring agricultural drought). 
  - However, more doubt can still raises: Whether the variable used to construct the index is a good set of variables. 
  - Does the index series capture the variables measures. Whether other ways of combining the variables results in difference in the index. 
  - Being able to answer these questions help to prove the robustness, extendibility and dimensionality of the index, as proposed by xxx as criteria for (selecting?) indexes.

@@

```{r width = "100%"}
knitr::include_graphics(here::here("figures/toy-example-spei.png"))
```

@@

## Pipeline steps for constructing indies

@@

>>>>

---
title: "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data"
format:
  tandf-pdf:
    keep-tex: true  
pdf-engine: pdflatex
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: 
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-3813-7155
  - name: Ursula Laa
    affiliation: 
      - name: University of Natural Resources and Life Sciences
        department: Institute of Statistics
        address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
        city: Vienna
        country: Austria
    orcid: 0000-0002-0249-6439
  - name: Nicolas Langrené
    affiliation: 
      - name: BNU-HKBU United International College
        department: Department of Mathematical Sciences
        address: 2000 Jintong Road, Tangjiawan, Zhuhai
        city: Zhuhai, Guangdong
        country: China
    orcid: 0000-0001-7601-4618
  - name: Patricia Menéndez
    affiliation:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0003-0701-6315
abstract: |
  - indexes, useful, quantify severity, early monitoring, 
  - A huge number of indexes have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
  - One of the reasons for this is the plenty of indexes are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
  - The paper describes a general pipeline framework to construct indexes from spatio-temporal data, 
  - This allows all the indexes to be constructed through a uniform data pipeline and different indexes to vary on the details of each step in the data pipeline and their orders. 
  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indexes into small building blocks shared by most of the indexes.
  - The framework will be demonstrated with drought indexes as examples, but appliable in general to environmental indexes constructed from multivariate spatio-temporal data
keywords:
  - indexes
  - data pipeline
  - software design
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: console
---

@@

While the [decision to make] is often multi-faceted, simple indexes tend to be favoured when it comes to communicating to stakeholders and public.

@@

While indexes are commonly used in operation and research, their compositions and structures can vary substantially. An index can be composed from a singe variable, such as the Standardized Precipitation Index (SPI) [@spi], or multiple variables, also known as composite indexes or composite indicators, like the Human Development Index [@hdi]. Some indexes can be expressed as a single equation, like Normalized Difference Vegetation Index (NDVI) [@ndvi], calculated as $(\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})$. In other cases, indexes themselves can be used as variables to construct indexes. For example, SPI is used to in many complex indexes, such as the U.S. Drought monitor and European Combined Drought Index (CDI) [@cdi]. While the index framework proposed in this paper applies to all the aforementioned indexes, our focus will be on studying complex indexes that combine multiple variables using statistical methods.

The initial plan aims to investigate how variable combinations can affect the indexes and whether we can use statistical methods to recommend better combination for decision making. However, when analysing specific index examples, we find the composition of indexes is much more complicated than originally expected. In addition to combining multiple variables, indexes can have a spatial and/or temporal component to capture the neighbouring effect in space and the prolonged effect in time. [Constructing indexes from combining variables with spatial and temporal component is further complicated by different treatments to process data from different sources such as satellite imageries, in-situ stations, and large physical models.]

@@

While numerous indexes are proposed by researchers, most well-known indexes only use one or two variables. More complex indexes tend to have difficulties to reach consensus given the uncertainty of their behaviors in different scenarios. Typical ways to evaluate new indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e.  crop yield for monitoring agricultural drought.

  - However, more doubt can still raises: Whether the variable used to construct the index is a good set of variables. 
  - Does the index series capture the variables measures. Whether other ways of combining the variables results in difference in the index. 
  - Being able to answer these questions help to prove the robustness, extendibility and dimensionality of the index, as proposed by xxx as criteria for (selecting?) indexes.

@@

```{r out.width = "100%"}
knitr::include_graphics(here::here("figures/toy-example-spei.png"))
```

@@

## Pipeline steps for constructing indexes

@@

  * align the spatial resolution and temporal frequency of data collected from different satellite products
  * obtain and clean variables from countries, potentially aggregate from regional data into country level
  * merge ground-based and satellite-based data

### 2023-07-09 tidyindex.tex
re-write the index definition section and shift of research direction in intro

\usepackage{orcidlink}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Demo arXiv template},
  pdfauthor={H. Sherry Zhang; Collaborators},
  pdfkeywords={indexes, data pipeline, software design},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Demo arXiv template}
\author{H. Sherry
Zhang$\textsuperscript{1}$~\orcidlink{0000-0002-7122-1463}, Collaborators$\textsuperscript{1}$}

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne, VIC, Australia
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

@@

\begin{itemize}
\tightlist
\item
  Indexes can be composed from single variables or multiple variables
  (composite index)
\item
  to make the definition even more complex is that indexes themselves
  are used with other variables to compose indexes (many drought indexes
  using an index called SPI).
\item
  Some indexes may be more simple than you think:

  \begin{itemize}
  \tightlist
  \item
    a single transformation (NDVI (Tucker 1979) takes two spectrum: red
    and near-infrared and calculates as
    \((\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})\))
  \item
    indexes sometimes are similar to features extracted from time series
    data: ETCCDI climate change indices (Zhang 2020): SU - number of
    summer days: annual count of days when TX \textgreater{} 25 degree
  \end{itemize}
\item
  The index pipeline and software implementation we propose are
  applicable to all indexes mentioned above, while the main focus of
  this paper is on indexes that combine multiple variables/indexes since
  more skills are required to diagnose those indexes to understand how
  they behave.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Difficult to reach consensus on composite/multivariate indexes.
\item
  Given the complexity of multivariate indexes as compared to univariate
  indexes, it raises more doubt to persuade people to use it.
\item
  Typical ways to evaluate new indexes is to compare it with the
  standard indexes (i.e.~SPI in drought index) or a response variable
  (i.e.~crop yield for monitoring agricultural drought).
\item
  However, more doubt can still raises: Whether the variable used to
  construct the index is a good set of variables.
\item
  Does the index series capture the variables measures. Whether other
  ways of combining the variables results in difference in the index.
\item
  Being able to answer these questions help to prove the robustness,
  extendibility and dimensionality of the index, as proposed by xxx as
  criteria for (selecting?) indexes.
\end{itemize}

@@

\includegraphics[width=10in,height=\textheight]{figures/toy-example-spei.png}

\hypertarget{pipeline-steps-for-constructing-indies}{%
\subsection{Pipeline steps for constructing
indies}\label{pipeline-steps-for-constructing-indies}}

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-tucker_red_1979}{}}%
Tucker, Compton J. 1979. {``Red and Photographic Infrared Linear
Combinations for Monitoring Vegetation.''} \emph{Remote Sensing of
Environment} 8 (2): 127--50.
\url{https://doi.org/10.1016/0034-4257(79)90013-0}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-climate-indexes}{}}%
Zhang, Xuebin. 2020. {``ETCCDI Climate Change Indices.''}
\url{http://etccdi.pacificclimate.org/index.shtml}.

>>>>

\usepackage{orcidlink}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Ursula Laa; Nicolas Langrené; Patricia Menéndez},
  pdfkeywords={indexes, data pipeline, software design},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Tidy Framework and Infrastructure to Systematically Assemble
Spatio-temporal Indexes from Multivariate Data}
\author{H. Sherry
Zhang$\textsuperscript{1}$~\orcidlink{0000-0002-7122-1463}, Dianne
Cook$\textsuperscript{1}$~\orcidlink{0000-0002-3813-7155}, Ursula
Laa$\textsuperscript{2}$~\orcidlink{0000-0002-0249-6439}, Nicolas
Langrené$\textsuperscript{3}$~\orcidlink{0000-0001-7601-4618}, Patricia
Menéndez$\textsuperscript{1}$~\orcidlink{0000-0003-0701-6315}}

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, boxrule=0pt, enhanced, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

While the {[}decision to make{]} is often multi-faceted, simple indexes
tend to be favoured when it comes to communicating to stakeholders and
public.

@@

While indexes are commonly used in operation and research, their
compositions and structures can vary substantially. An index can be
composed from a singe variable, such as the Standardized Precipitation
Index (SPI) (McKee, Doesken, and Kleist 1993), or multiple variables,
also known as composite indexes or composite indicators, like the Human
Development Index (United Nations Development Programme 2022). Some
indexes can be expressed as a single equation, like Normalized
Difference Vegetation Index (NDVI) (Tucker 1979), calculated as
\((\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})\). In other cases,
indexes themselves can be used as variables to construct indexes. For
example, SPI is used to in many complex indexes, such as the U.S.
Drought monitor and European Combined Drought Index (CDI)
(Sepulcre-Canto et al. 2012). While the index framework proposed in this
paper applies to all the aforementioned indexes, our focus will be on
studying complex indexes that combine multiple variables using
statistical methods.

The initial plan aims to investigate how variable combinations can
affect the indexes and whether we can use statistical methods to
recommend better combination for decision making. However, when
analysing specific index examples, we find the composition of indexes is
much more complicated than originally expected. In addition to combining
multiple variables, indexes can have a spatial and/or temporal component
to capture the neighbouring effect in space and the prolonged effect in
time. {[}Constructing indexes from combining variables with spatial and
temporal component is further complicated by different treatments to
process data from different sources such as satellite imageries, in-situ
stations, and large physical models.{]}

@@

While numerous indexes are proposed by researchers, most well-known
indexes only use one or two variables. More complex indexes tend to have
difficulties to reach consensus given the uncertainty of their behaviors
in different scenarios. Typical ways to evaluate new indexes is to
compare it with the standard/ well-known indexes, such as SPI or SPEI
for drought indexes, or a response variable relevant for the monitoring
purpose, i.e.~crop yield for monitoring agricultural drought.

\begin{itemize}
\tightlist
\item
  However, more doubt can still raises: Whether the variable used to
  construct the index is a good set of variables.
\item
  Does the index series capture the variables measures. Whether other
  ways of combining the variables results in difference in the index.
\item
  Being able to answer these questions help to prove the robustness,
  extendibility and dimensionality of the index, as proposed by xxx as
  criteria for (selecting?) indexes.
\end{itemize}

@@

\includegraphics[width=1\textwidth,height=\textheight]{figures/toy-example-spei.png}

\hypertarget{pipeline-steps-for-constructing-indexes}{%
\subsection{Pipeline steps for constructing
indexes}\label{pipeline-steps-for-constructing-indexes}}

@@

\begin{itemize}
\tightlist
\item
  align the spatial resolution and temporal frequency of data collected
  from different satellite products
\item
  obtain and clean variables from countries, potentially aggregate from
  regional data into country level
\item
  merge ground-based and satellite-based data
\end{itemize}

@@

\leavevmode\vadjust pre{\hypertarget{ref-cdi}{}}%
Sepulcre-Canto, G., S. Horion, A. Singleton, H. Carrao, and J. Vogt.
2012. {``Development of a {Combined} {Drought} {Indicator} to Detect
Agricultural Drought in {Europe}.''} \emph{Natural Hazards and Earth
System Sciences} 12 (11): 3519--31.
\url{https://doi.org/10.5194/nhess-12-3519-2012}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ndvi}{}}%
Tucker, Compton J. 1979. {``Red and Photographic Infrared Linear
Combinations for Monitoring Vegetation.''} \emph{Remote Sensing of
Environment} 8 (2): 127--50.
\url{https://doi.org/10.1016/0034-4257(79)90013-0}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
---------. 2022. {``Human Development Report 2021-22.''} New York.
\url{http://report.hdr.undp.org}.

@@
